Flabby and injective objects in toposes

### 2022-02-15 paper-flabby-objects.tex
Started on enough injectives without LEM

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}

@@

\begin{proof}\begin{enumerate}
\item One can check that, for instance, the full powerset~$\P(X)$ and the set of
subterminals~$\P_{\leq 1}(X)$ are each injective.
\item This statement follows from general abstract nonsense, since the forgetful functor from modules to sets possesses a monomorphism-preserving left
adjoint. More explicitly, let~$I$ be an injective~$R$-module, let~$i : A \to B$ be an injective map between sets and
let~$f : A \to I$ be an arbitrary map. Then the induced map~$R\langle A \rangle
\to R\langle B \rangle$ between free modules is also injective, the given
map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
extension~$R\langle B \rangle \to I$ induces an extension~$B \to I$ of~$f$.
\item One verifies that any abelian group embeds into a divisible
abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
The result for modules over arbitrary rings then follows purely
formally, since the functor~$A \mapsto \Hom(R,A)$ from abelian groups
to~$R$-modules has a left exact left adjoint with monic unit. \qedhere
\end{enumerate}\end{proof}

We note in passing that Remark~\ref{rem:enough-injective-modules-without-lem}
below explains how to verify Proposition~\ref{prop:basics-injective}(3) also in
the absence of the law of excluded middle (but still requiring Zorn's lemma).

@@

We currently believe that it is not possible to give a constructive account of
a global cohomology functor which would associate to any sheaf of modules its
cohomology. However, it should be possible to do so for a restricted class of
sheaves, while still preserving the good formal properties expected from
derived functors. To this end, a mix of the approaches using
pointwise Kan extensions, as cogently argued for by Emily
Riehl~\cite[Chapter~2]{riehl:cathtpy}, and ind-objects as presented in the Stacks
Project~\cite[Tag~05S7]{stacks-project} seems promising.

This framework is sufficiently flexible to not demand derived functors to be
defined everywhere. Rather, they will be defined just for those objects for
which we happen to have a suitable resolution. Only classically, by using
injective or flabby sheaves, can we pretend that every object has such a
resolution. We hope to report on details in future work.

@@

@@

TO DO:
* enough injectives employing only Zorn
* first names?

>>>>

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}

@@

\begin{proof}\begin{enumerate}
\item One can check that, for instance, the full powerset~$\P(X)$ and the set of
subterminals~$\P_{\leq 1}(X)$ are each injective.
\item This statement follows from general abstract nonsense, since the forgetful functor from modules to sets possesses a monomorphism-preserving left
adjoint. More explicitly, let~$I$ be an injective~$R$-module, let~$i : A \to B$ be an injective map between sets and
let~$f : A \to I$ be an arbitrary map. Then the induced map~$R\langle A \rangle
\to R\langle B \rangle$ between free modules is also injective, the given
map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
extension~$R\langle B \rangle \to I$ induces an extension~$B \to I$ of~$f$.
\item One verifies that any abelian group embeds into a divisible
abelian group. By Baer's criterion (which in this form requires the axiom of choice), divisible abelian groups are injective.
The result for modules over arbitrary rings then follows purely
formally, since the functor~$A \mapsto \Hom(R,A)$ from abelian groups
to~$R$-modules has a left exact left adjoint with monic unit. \qedhere
\end{enumerate}\end{proof}

We note in passing that the multi-step technique of the proof of
Theorem~\ref{thm:enough-flabby-modules} below can be used to verify
Proposition~\ref{prop:basics-injective}(3) also in the absence of the law of
excluded middle (but still requiring Zorn's lemma). Since this observation is of no
further import for the purposes of this text, details are postponed to
Section~\ref{sect:enough-injective-modules}.

@@

We currently believe that it is not possible to give a constructive account of
a global cohomology functor which would associate to any sheaf of modules its
cohomology. However, it should be possible to do so for a restricted class of
sheaves, while still preserving the good formal properties expected from
derived functors. To this end, a mix of the approaches using
pointwise Kan extensions, as cogently argued for by Emily
Riehl~\cite[Chapter~2]{riehl:cathtpy} (see also~\cite{hinich:what-is,maltsiniotis:adjunction}), and ind-objects as presented in the Stacks
Project~\cite[Tag~05S7]{stacks-project} seems promising.

This framework is sufficiently flexible to not demand derived functors to be
defined everywhere. Rather, they will be defined just for those objects for
which we happen to have a suitable resolution. Only classically, by using
injective or flabby sheaves, can we pretend that every object has such a
resolution. We hope to report on details in future work. Can perhaps even the
rare cases where injective resolutions exceptionally do have some link to
computations, as on projective spaces~\cite{huang:cohomology}, be given a
proper constructive and predicative home?

@@

\appendix

% \begin{document}

\section{On the existence of enough injective modules}
\label{sect:enough-injective-modules}

\begin{defn}\begin{enumerate}
\item An abelian group~$I$ is \emph{divisible} if and only if for every
element~$x \in I$ and every natural number~$n \geq 1$, there is an element~$y \in I$
such that~$ny = x$.
\item An abelian group~$I$ \emph{satisfies the Baer criterion} if and only if,
for every ideal~$\aaa \subseteq \ZZ$, every ($\ZZ$-)linear map~$\aaa \to I$
admits an extension along the inclusion to a linear map~$\ZZ \to I$.
\end{enumerate}
\end{defn}

\begin{prop}
\begin{enumerate}
\item Assuming the law of excluded middle, divisible abelian groups satisfy
the Baer criterion.
\item Assuming Zorn's lemma, abelian groups which satisfy the Baer criterion
are injective.
\end{enumerate}
\end{prop}

\begin{proof}Assuming the law of excluded middle, the only ideals of~$\ZZ$ are
the principal ideals~$(n)$ where~$n \geq 0$. For~$n = 0$ existence of
extensions is trivial and for~$n > 0$ existence of extensions follow from
divisibility.

Assuming Zorn's lemma, let~$I$ be an abelian group satisfying the Baer
criterion, let~$i : A \to B$ be a linear injection and let~$f : A \to I$ be a
linear map. The poset of partial extensions contains suprema of chains and
hence a maximal partial extension~$f_0 : A_0 \to I$ with~$A \subseteq A_0
\subseteq B$. To verify that~$A_0 = B$, let an element~$x \in B$ be given.
By the Baer condition, the linear map~$g : (A_0 : x) \to I$ defined on the
ideal~$(A_0 : x) = \{ n \in \ZZ \,|\, nx \in A_0 \} \subseteq \ZZ$ given by~$g(n) =
f_0(nx)$ can be extended to a linear map~$\overline{g} : \ZZ \to I$.
Hence~$f_0$ can be extended to the map~$A_0 + (x) \to I$ given by~$u + nx
\mapsto f_0(u) + \overline{g}(n)$. By maximality~$A_0 + (x) = A_0$, hence~$x \in A_0$.
\end{proof}

\begin{prop}
\begin{enumerate}
\item Any abelian group embeds into a divisible abelian group.
\item Any abelian group embeds into an abelian group satisfying the Baer
condition.
\end{enumerate}
\end{prop}

\begin{proof}For the first part, an abelian group~$A$ embeds into the
divisible group~$\QQ\langle A \rangle / K$, where~$\QQ\langle A \rangle$ is the
underlying abelian group of the free~$\QQ$-vector space on~$A$ and~$K$ is the
kernel of the canonical map~$\ZZ\langle A \rangle \to A$.

The second part is more involved and requires, similar to the proof of
Theorem~\ref{thm:enough-flabby-modules}, a strengthening of the Baer condition
where the required extensions are explicitly given. As a first step, let~$A$ be
an abelian group which is a sheaf for a modal operator~$\nabla$ such
that~$\nabla(\varphi \vee (\varphi \Rightarrow \nabla\bot))$. Let~$i : \aaa
\hookrightarrow \ZZ$ be the inclusion of an ideal of~$\ZZ$ and let~$f : \aaa
\to A$ be a linear map. Applying the sheafification functor~$(\cdot)^{++}$, we
have the following commutative diagram:
\[ \xymatrix{
  \aaa \ar@{^{(}->}[r]^i\ar[d]\ar@/_2pc/[dd]_f & \ZZ \ar[d] \\
  \aaa^{++} \ar@{^{(}->}[r]^{i^{++}}\ar@{-->}[d] & \ZZ^{++}\ar@{-->}@/^/[ld]^{f'} \\
  A
} \]
The given map~$f$ factors (uniquely) over the map~$\aaa \to \aaa^{++}$ since~$A$ is a
sheaf. The map~$i^{++}$ is still injective as sheafification is exact. The two
top vertical maps need not be injective, as~$\ZZ$ need not be separated
for~$\nabla$; but this issue XXX
\end{proof}

\begin{prop}Assuming Zorn's lemma:
\begin{enumerate}
\item Any abelian group embeds into an injective abelian group.
\item Any module embeds into an injective module.
\item XXX Grothendieck topos?
\end{enumerate}
\end{prop}

@@

TO DO:
* enough injectives employing only Zorn
* first names?
* cite more from Barakat and friends?
